---
title: "Getting Started with NHANES Data"
format: html
editor: visual
vignette: >
  %\VignetteIndexEntry{Getting Started with NHANES Data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Why NHANES Data Matters (And Why It's Been Frustrating to Use)

The National Health and Nutrition Examination Survey (NHANES) is one of the most comprehensive public health datasets available. With data on tens of thousands of Americans spanning over two decades, it's an invaluable resource for understanding trends in health, nutrition, disease, and demographics.

If you've tried to work with NHANES data before, you've likely encountered two major pain points:

1. **CDC server reliability**: The CDC's data servers can be slow, unresponsive, or down entirely. This makes reproducible research difficult when you can't count on data availability.  
2. **Cycle suffix confusion**: The CDC publishes data in two-year cycles, each with a different letter suffix. Want demographics data? You'll need to work with `DEMO`, `DEMO_B`, `DEMO_C`, all the way through `DEMO_L`. Finding the right documentation becomes a scavenger hunt, and combining cycles requires careful data wrangling.  

The **nhanesdata** package solves both problems:  

- **Reliable access**: All datasets are hosted on Cloudflare R2 with a public URL, giving you fast, dependable access anytime.  
- **Simplified naming**: We've already merged all cycles for you. Just ask for `demo` and you get all demographics data from 1999-2021, with a `year` column to track which cycle each observation belongs to.  

Let's get you from zero to your first NHANES analysis in the next 10 minutes.

## Installation

Install the package from GitHub:

```{r}
#| label: install
#| eval: false

# install.packages("devtools")
devtools::install_github("kyleGrealis/nhanesdata")
```

Then load the package:

```{r}
#| label: load-packages
#| eval: false

library(nhanesdata)
library(dplyr)    # for data manipulation
library(ggplot2)  # for visualization
```

## Your First Dataset in 30 Seconds

Let's load the demographics dataset, which contains age, gender, race/ethnicity, income, and other demographic variables:

```{r}
#| label: load-demo
#| eval: false

demo <- read_nhanes('demo')

# Take a look at what we have
glimpse(demo)
```

Notice the `year` and `seqn` columns:

- `year`: The survey cycle start year (1999, 2001, 2003, ..., 2017, 2021)  
- `seqn`: The respondent sequence number (unique within a cycle, used for joining datasets)  

With this one simple call, you now have over 100,000 observations spanning 11 survey cycles. No server timeouts, no wrestling with cycle suffixes.

## Quick Analysis: Age Distribution Over Time

Let's examine how the age distribution of NHANES participants has evolved:

```{r}
#| label: age-analysis
#| eval: false

demo %>%
  filter(!is.na(ridageyr)) %>%
  ggplot(aes(x = ridageyr)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  facet_wrap(~year, ncol = 4) +
  labs(
    title = "NHANES Age Distribution by Survey Cycle",
    x = "Age (years)",
    y = "Count"
  ) +
  theme_minimal()
```

In just a few lines, you've visualized trends across more than 20 years of public health data. That's the power of simplified access.

## Understanding CDC Dataset Names (Important!)

Here's where the CDC's naming system can trip you up. When you use `read_nhanes('demo')`, you're getting data that the CDC publishes across multiple tables:

| **CDC Table Name** | **Survey Years** | **What nhanesdata Does**        |
|--------------------|------------------|---------------------------------|
| DEMO               | 1999-2000        | Merges all of these into a      |
| DEMO_B             | 2001-2002        | single dataset called `demo`,   |
| DEMO_C             | 2003-2004        | adds a `year` column, and       |
| ...                | ...              | harmonizes data types across    |
| DEMO_L             | 2021-2022        | cycles so you can analyze them  |
|                    |                  | together without any extra work |

**Why does this matter?**

When you need to look up what a variable means, you'll still need to reference the CDC's documentation, which uses the cycle-specific names. For example, if you're working with 2015-2016 data, you'll want the `DEMO_I` codebook.

The `get_url()` function helps with this:

```{r}
#| label: get-url
#| eval: false

# Get documentation for the original DEMO table (1999-2000)
get_url('DEMO')

# Get documentation for DEMO_I (2015-2016)
get_url('DEMO_I')
```

This will return the full URL to the CDC's documentation page where you can find variable descriptions, value codes, and survey methodology notes.

## Combining Multiple Datasets

The real power of NHANES comes from linking datasets together. Let's combine demographics with body measures (height, weight, BMI) to look at BMI trends:

```{r}
#| label: bmi-analysis
#| eval: false

# Load body measures data
bmx <- read_nhanes('bmx')

# Join with demographics
bmi_data <- demo %>%
  select(seqn, year, ridageyr, riagendr) %>%
  inner_join(
    bmx %>% select(seqn, year, bmxbmi),
    by = c("seqn", "year")
  )

# Analyze BMI trends for adults by gender
bmi_data %>%
  filter(
    ridageyr >= 20,           # adults only
    !is.na(bmxbmi),           # remove missing BMI
    !is.na(riagendr)          # remove missing gender
  ) %>%
  group_by(year, riagendr) %>%
  summarize(
    mean_bmi = mean(bmxbmi, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = year, y = mean_bmi, color = factor(riagendr))) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  scale_color_manual(
    values = c("1" = "steelblue", "2" = "coral"),
    labels = c("1" = "Male", "2" = "Female"),
    name = "Gender"
  ) +
  labs(
    title = "Mean BMI Trends Among U.S. Adults",
    subtitle = "NHANES 1999-2021",
    x = "Survey Year",
    y = "Mean BMI"
  ) +
  theme_minimal()
```

**Key joining principle**: Always join on both `seqn` AND `year`. The same person can appear in multiple survey cycles (though this is uncommon), and the same sequence number can be reused across cycles.

## Finding Variables: Working Backwards from Your Question

Let's say you want to study hypertension. You know NHANES collects blood pressure data, but which dataset contains it?

### Method 1: Search by Term

Use `term_search()` to find variables related to a topic:

```{r}
#| label: term-search
#| eval: false

# Find variables related to blood pressure
bp_vars <- term_search("blood pressure")
head(bp_vars)
```

This returns a dataset showing:

- `Variable.Name`: The variable code (e.g., `BPXSY1` for systolic blood pressure)  
- `Data.File.Name`: Which CDC table it's in (e.g., `BPX_J`)  
- `Data.File.Description`: What the table contains  
- `Begin.Year`: When this data was collected  

From this, you can see that blood pressure data lives in the `BPX` tables. In nhanesdata, that means you want `read_nhanes('bpx')`.

### Method 2: Search by Variable Name

If you know the specific variable you need, use `var_search()`:

```{r}
#| label: var-search
#| eval: false

# Find all occurrences of the BPXSY1 variable
var_search("BPXSY1")
```

This shows you every cycle where that variable appears, which is helpful for understanding data availability over time.

### Method 3: Verify Variables in Your Loaded Data

If you've already loaded a dataset and want to confirm it contains the variable you need:

```{r}
#| label: check-var
#| eval: false

# Load a dataset
bmx <- read_nhanes('bmx')

# Check if it contains the height variable
'bmxht' %in% names(bmx)  # TRUE

# Or see all available columns
names(bmx)
```

This approach helps you verify that variables from the CDC documentation are actually present in the merged dataset.

## A Complete Example: Blood Pressure by Age Group

Let's put it all together with a realistic analysis examining how blood pressure varies by age:

```{r}
#| label: complete-example
#| eval: false

# Load required datasets
demo <- read_nhanes('demo')
bpx <- read_nhanes('bpx')

# Combine and analyze
bp_analysis <- demo %>%
  filter(ridageyr >= 18) %>%  # adults only
  select(seqn, year, ridageyr, riagendr, ridreth1) %>%
  inner_join(
    bpx %>% select(seqn, year, bpxsy1, bpxdi1),
    by = c("seqn", "year")
  ) %>%
  filter(
    !is.na(bpxsy1),     # valid systolic BP
    !is.na(bpxdi1),     # valid diastolic BP
    bpxsy1 > 0,         # exclude special codes
    bpxdi1 > 0
  ) %>%
  mutate(
    age_group = cut(
      ridageyr,
      breaks = c(18, 30, 40, 50, 60, 70, 80, Inf),
      labels = c("18-29", "30-39", "40-49", "50-59", "60-69", "70-79", "80+"),
      right = FALSE
    )
  )

# Calculate mean BP by age group
bp_summary <- bp_analysis %>%
  group_by(age_group) %>%
  summarize(
    n = n(),
    mean_systolic = mean(bpxsy1),
    mean_diastolic = mean(bpxdi1),
    .groups = "drop"
  )

# Visualize
bp_summary %>%
  ggplot(aes(x = age_group)) +
  geom_col(aes(y = mean_systolic), fill = "coral", alpha = 0.7) +
  geom_col(aes(y = mean_diastolic), fill = "steelblue", alpha = 0.7) +
  labs(
    title = "Blood Pressure Increases with Age",
    subtitle = "Mean systolic (coral) and diastolic (blue) BP by age group",
    x = "Age Group",
    y = "Blood Pressure (mmHg)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

This analysis combines data from over 100,000 individuals across multiple survey cycles to reveal the well-known pattern of increasing blood pressure with age. You did this without dealing with a single CDC server timeout or cycle suffix.

## Summary: What You've Learned

In this vignette, you've learned how to:

1. Load NHANES datasets instantly with `read_nhanes()`
2. Understand the relationship between CDC cycle-specific tables and nhanesdata's merged datasets
3. Join multiple datasets using `seqn` and `year`
4. Find variables using `term_search()` and `var_search()`
5. Access CDC documentation with `get_url()`
6. Conduct real analyses combining demographic, anthropometric, and examination data

The nhanesdata package removes the traditional friction from working with NHANES data. You can now focus on answering your research questions instead of fighting with data infrastructure.

## Next Steps

Ready to dive deeper? Here are some ideas for your own analyses:  

- Examine trends in diabetes prevalence using the `diq` (diabetes questionnaire) dataset  
- Analyze changes in dietary patterns with `dr1tot` and `dr2tot` (dietary interview data)  
- Study relationships between physical activity (`paq`) and health outcomes  
- Explore prescription medication use patterns with `rxq_rx` data  

All of these datasets are available through `read_nhanes()`. Use the search functions to explore what's available, and don't hesitate to combine multiple datasets to answer complex questions.

The data is waiting for you. Happy analyzing!
