---
title: "Introduction to nhanesdata"
format: html
editor: visual
vignette: >
  %\VignetteIndexEntry{Introduction to nhanesdata}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

This vignette shows you how to search for NHANES variables and load pre-processed datasets using the nhanesdata package.

## Acknowledgments

This package builds on the excellent [nhanesA](https://cran.r-project.org/package=nhanesA) package, which provides the foundation for accessing NHANES data through R. We're grateful for the nhanesA developers' work in making CDC data accessible.

The nhanesdata package extends nhanesA by providing:  
- Pre-processed datasets for fast access  
- Automatic harmonization across survey cycles (1999-2021)  
- Convenient search and discovery functions  

## Setup

First, load the `nhanesdata` package to get access to all its functions.

```{r}
#| label: setup
#| message: false
#| eval: false

library(nhanesdata)
library(dplyr)
```

## Searching for Data

The package includes several helper functions to find specific NHANES data. **All functions are case-insensitive!**
  
*   `term_search()`: Search for a term across variable names and descriptions  
*   `var_search()`: Find the datasets a specific variable belongs to  
*   `get_url()`: Get a direct link to the CDC's documentation for a data table  

For example, let's find data related to "caffeine":

```{r}
#| label: search-example
#| eval: false

# Search functions are case-insensitive
term_search("caffeine")
term_search("CAFFEINE")  # Same results!
```

## Reading Pre-Processed Data (Recommended)

The primary way to use this package is with `read_nhanes()`, which loads pre-processed NHANES datasets from cloud storage.

**Why "read_nhanes"?** The datasets are hosted on Cloudflare R2 (cloud object storage) for fast, reliable worldwide access. No credentials needed—the function downloads and reads the data in one step.

### Benefits:  
- **Fast downloads** from cloud storage (hosted on Cloudflare R2)  
- **All survey cycles (1999-2021)** already merged  
- **Type inconsistencies** already resolved  
- **Updated quarterly** via automated workflows  
- **Case-insensitive dataset names** - use 'demo', 'DEMO', or 'Demo' interchangeably!  

### Example:

```{r}
#| label: read-nhanes-example
#| eval: false

# All case variations work identically:
trigly <- read_nhanes('trigly')    # Lowercase (easiest to type)
demo <- read_nhanes('DEMO')        # Uppercase (matches CDC docs)
bpx <- read_nhanes('Bpx')          # Mixed case (because you can!)

head(trigly)
```

**Note:** Each dataset includes a `year` column that identifies which survey cycle each row came from.

## Important: 2019-2020 Survey Cycle

The 2019-2020 survey cycle (suffix K) is **not included** in the datasets provided by this package.

The COVID-19 pandemic disrupted data collection during this cycle, and the CDC determined that the data quality was insufficient for standard NHANES analysis. The CDC combined partial 2019-2020 data with 2021-2023 data and released it as a special pre-pandemic dataset.

**Reference:** [CDC NHANES 2019-2020 Data Documentation](https://wwwn.cdc.gov/nchs/nhanes/continuousnhanes/default.aspx?BeginYear=2019)

**What this means:** Datasets jump from suffix J (2017-2018) to suffix L (2021-2023), skipping K entirely.

## Working with Multiple Datasets

A common workflow is to load and join multiple related datasets:

```{r}
#| label: multiple-datasets
#| eval: false

library(nhanesdata)
library(dplyr)

# Dataset names are case-insensitive - mix freely!
demo <- read_nhanes('demo')        # Lowercase
bpx <- read_nhanes('BPX')          # Uppercase
bmx <- read_nhanes('Bmx')          # Mixed case

# Join by participant ID (seqn) and survey year
analysis_data <- demo |>
  inner_join(bpx, by = c('seqn', 'year')) |>
  inner_join(bmx, by = c('seqn', 'year')) |>
  select(year, seqn, ridageyr, riagendr, bpxsy1, bmxbmi)
```

The `year` column ensures you only join observations from the same survey cycle.

## Filtering by Survey Year

Since datasets include all cycles, you can filter to specific time periods:

```{r}
#| label: filter-years
#| eval: false

# Load demographics
demo <- read_nhanes('demo')

# Filter to recent cycles only (2015 and later)
recent_demo <- demo |>
  filter(year >= 2015)

# Compare across time periods
demo |>
  mutate(
    period = case_when(
      year < 2010 ~ '1999-2009',
      year < 2020 ~ '2010-2019',
      TRUE ~ '2020+'
    )
  ) |>
  group_by(period) |>
  summarise(n_participants = n())
```
