---
title: "Available NHANES Datasets"
format: html
vignette: >
  %\VignetteIndexEntry{Available NHANES Datasets}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false

library(dplyr)
library(reactable)

knitr::opts_chunk$set(
  collapse = TRUE,
  eval = FALSE, # Default: don't evaluate most chunks
  comment = "#>"
)

config <- yaml::read_yaml(
  system.file("extdata", "datasets.yml", package = "nhanesdata")
)

catalog <- do.call(rbind, lapply(config$datasets, function(x) {
  data.frame(
    Dataset = toupper(x$name),
    Description = x$description,
    Category = tools::toTitleCase(x$category),
    stringsAsFactors = FALSE
  )
}))
```

## Available Datasets

This package provides `r nrow(catalog)` NHANES datasets, automatically updated quarterly with data from 1999-2023 (excluding the 2019-2020 cycle).

### Quick Start
```r
library(nhanesdata)

# Load demographics data
demo <- read_nhanes('demo')

# Search for variables
term_search('blood pressure')
```


### Categories

**Questionnaire/Interview Tables** - Self-reported data from participant interviews

**Examination Tables** - Physical measurements and laboratory results

```{r}
#| label: reactable-of-datasets
#| echo: false
#| eval: true

catalog |>
  arrange(Dataset) |>
  reactable::reactable(
    searchable = TRUE
  )
```

### Notes

- All datasets span multiple survey cycles (1999-2023)
- Each includes `year` and `seqn` columns for merging
- Data types are harmonized across cycles
- Variable names match CDC documentation

For detailed variable information, use `term_search()` or visit the [CDC NHANES website](https://wwwn.cdc.gov/nchs/nhanes/).

::: callout-warning
CDC may change data periodically. The data was aggregated as best as possible to reconcile
variable types that changed across cycles. **ALWAYS** reference the CDC documentation
with `nhanesdata::get_url(dataset)`!

+ See [get_url](.docs/reference/get_url.html) documentation.
:::
